Outline flowCore paper

### 2008-07-22 paper/manuscript.tex
some updates to the paper
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@32839 bc3139a8-67e5-0310-9ffc-ced21a209358

The computational tools we have developed are distributed in the R
software language (www.r-project.org) as the Bioconductor
(www.bionductor.org) package flowCore. flowCore is a freely available,
highly functional and extensible FCM data analysis platform that
enables researchers to efficiently handle FC-HCS data and encourages
open development of tools for their coherent analysis. Among the
issues addressed by flowCore are computationally efficient data
structures and a range of specialized methods for compensation,
transformation, and gating.  flowCore runs on Windows, Mac OS X and
Linux/Unix operating systems. Conceptually, the underlying data
structures that build the core of our software can be implemented in
any programming language, and we chose the R language for its obvious
advantages in statistical programming. In this paper, we describe the
functionality and methods underlying flowCore, and provide examples of
their use.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Gates} \\
\hline
rectangleGate & n-dinensional rectangular regions \\
quadGate & quadrant regions in two dimensions \\
polygonGate & polygonal regions in two dimensions \\
polytopeGate & generalization of polygon in n dimensions \\
ellipsoidGate & n-dimesnional ellipsoid region \\
\hline
\multicolumn{2}{|c|}{Filters} \\
\hline
sampleFilter & random subsampling\\
expressionFilter & filter by resuts of a boolean expression \\
kmeansFilter & based on K-means clustering \\
norm2Filter & based on bivariate normal distribution \\
curv1Filter & based on local density regions in 1D \\
curv2Filter & based on local density regions in 2D \\
timeFilter & based on abnormal data acquisition over time \\
\hline
filterSet & gating strategy \\
\hline
\end{tabular}
\caption{\label{table2}Filter and gate classes implemented in flowCore.}
\end{center}
\end{table}

\subsection*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of FCM data analysis
%(Figure \ref{fig2:flowWheel}) 
. The flowViz package provides
sophisticated data visualization tools, that make use of multivariate
trellis plotting. These functions can be used to quickly generate
customized plots for extended cytometry data sets for both direct data
inspection and quality control. The flowQ package offers more advanced
quality assurance (QA) methodology and a framework to create
interactive web-based reports of QA results. Most of flowUtil's
content deals with data import and output from and to flowCore and the
flow-cytometry specific standard markup languages. Finally, flowStats
provides statistical methods that are relevant in the context of flow
cytometry data analysis.

%\begin{figure}
%\centering
%\includegraphics[width=0.6\textwidth]{Figure2-flowWheel-version4.jpg}
%\caption{\label{fig2:flowWheel}{flowCore and assoziated Bioconductor
%    packages: dedicated packages have been developed that utilize
%    flowCoreœôòùs data structures. flowViz is a data visualization
%    package; flowQ provides data quality assessment methods; flowUtils
%    contains interfaces for flow data, including xml parsers:
%    flowClust provides automated gating by model-based clustering.}}
%\end{figure}

\section*{Application}
The flowCore package has successfully been applied in the analysis of
several datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
several other Bioconductor packages are making use of its code
base. As an example, \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using the
flowCore data model and infrastructure and implemented this in the
Bioconductor package flowClust. Another package (plateCore) providing
more specialized support for experiments conducted on microtitre
plates is under development. The documentation of all of flowCore's
features is beyond the scope of this publication, however additional
documentation and users guide information with programmatic examples
is available <here> . However, in the Supplementary Material to this
article we use a single example to demonstrate some of the software's
hallmarks, including the concept of data-driven automated gating, the
integration of existing software into the framework, and the
generation of publication-quality graphics for data visualization.

%% not sure about this next section vs. the use of Suuplementary material. Would it be useful/informative to show the code required to generate the plot? Otherwise I'm not sold that the next paragraph adds anything.

%% I am not sure about the use of Supplementary Material - flowCore and all the packages mentioned here 
%% already have at least 1 vignette with examples and code

A very basic matrix of scatterplots is shown in Figure \ref{xyplot},
where each panel in the matrix represents the data of one individual
patient. Instead of plotting individual points, the local bivariate
density of the data is shown as a false color image. For each panel,
the outlines of the regions identified in the automated gating
operation described below are added to the plot. The input to the
plotting function is either a single flowFrame or a complete flowSet
object, and all available metadata can be accessed to control the
appearance and the composition of the plot. Similarly, gate objects
are directly passed on to the plotting function and are included in
the plot if possible. Besides the very basic scatterplot shown here,
there is a wealth of additional data visualizations available in the
flowViz package\cite{sarkar2008ufv}, all making use of flowCore's data
models allowing to easily arrange the data according to numerous
experimental factors. Furthermore, the design and the API of the
visualization software itself is very generic, and the user can
readily extend its capabilities by providing self-defined plotting
functions.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
Scatterplot matrix of a single FlowSet with outlines of the gating
regions identified by an automated gating operation.}
\end{figure}

Static gating for all samples in a high-throughput FCM experiment is
often not an option, since the measured variables tend to vary between
different treatments or over time. Automated or data-driven gating
tries to estimate the gating regions from the underlying data, thus
providing a fast objective solution to the analysis of potentially
very large and diverse data sets \cite{lo2008agf}. One of the
automated gating methods implemented in flowCore is based on
identifying areas of significant curvature in a kernel density
estimate of the data \citep{wand2008}. Assuming that the regions of
interest are of high density, the software is able to reliably detect
them in either the one or the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it. The modular design of R,
Bioconductor and flowCore allows to easily integrate these existing
solutions into our framework. In this example, we directly use R code
from the feature package developed by \cite{wand2008}. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

\section*{Conclusions}

>>>>

The computational tools we have developed are distributed in the R
software language \citep{Rmain} as the Bioconductor
\citep{BIOC} package flowCore. flowCore is a freely available,
highly functional and extensible FCM data analysis platform that
enables researchers to efficiently handle FC-HCS data and encourages
open development of tools for their coherent analysis. Among the
issues addressed by flowCore are computationally efficient data
structures and a range of specialized methods for compensation,
transformation, and gating.  flowCore runs on Windows, Mac OS X and
Linux/Unix operating systems. Conceptually, the underlying data
structures that build the core of our software can be implemented in
any programming language, and we chose the R language for its obvious
advantages in statistical programming. In this paper, we describe the
functionality and methods underlying flowCore, and provide examples of
their use.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l|l|}
\hline
\multicolumn{2}{|c|}{Gates} \\
\hline
rectangleGate & n-dinensional rectangular regions \\
quadGate & quadrant regions in two dimensions \\
polygonGate & polygonal regions in two dimensions \\
polytopeGate & generalization of polygon in n dimensions \\
ellipsoidGate & n-dimesnional ellipsoid region \\
\hline
\multicolumn{2}{|c|}{Filters} \\
\hline
sampleFilter & random subsampling\\
expressionFilter & filter by resuts of a boolean expression \\
kmeansFilter & based on K-means clustering \\
norm2Filter & based on bivariate normal distribution \\
curv1Filter & based on local density regions in 1D \\
curv2Filter & based on local density regions in 2D \\
timeFilter & based on abnormal data acquisition over time \\
\hline
filterSet & gating strategies \\
\hline
\end{tabular}
\caption{\label{table2}Filter and gate classes implemented in flowCore.}
\end{center}
\end{table}

\subsection*{Related flow packages}
In addition to the flowCore package that offers basic infrastructure,
we have implemented a range of additional Bioconductor packages that
are dedicated to more specific tasks of FCM data analysis. The flowViz
package \citep{sarkar2008ufv} provides sophisticated data
visualization tools, that make use of multivariate trellis
plotting. These functions can be used to quickly generate customized
plots for extended cytometry data sets for both direct data inspection
and quality control. The objects annotation information can be used to
arange the layout and composition of the plots. Furthermore, the
design and the API of the visualization software is very generic, and
users can readily extend its capabilities by providing self-defined
plotting functions.  The flowQ package offers more advanced quality
assurance (QA) methodology and a framework to create interactive
web-based reports of QA results. Most of flowUtil's content deals with
data import and output from and to flowCore and the flow-cytometry
specific standard markup languages. Finally, flowStats provides more
elaborate statistical methods that are relevant in the context of flow
cytometry data analysis.

\section*{Results}
The flowCore package has been successfully applied in the analysis of
several datasets \citep{gasparetto2004ice,brinkman2007hcf}, and
several other Bioconductor packages are making use of its code
base. As an example, \cite{lo2008agf} have recently developed an
automatic gating approach via robust model-based clustering using
flowCore's data model and infrastructure and implemented this in the
Bioconductor package flowClust. Another package, plateCore, providing
more specialized support for experiments conducted on microtitre
plates is under development.

A complete documentation of the flowCore software is way beyond the
scope of this publication. Much more comprehensive documentation and
users guide information with programmatic examples is available online
(\url{http://bioconductor.org/packages/2.2/bioc/html/flowCore.html})
and as part of the package distribution. Here, we only want to briefly
exemplify some of the software's key features, that is, the coherent
treatment of all samples in a potentially large expreriment, the
concept of data-driven automated gating, the integration of existing
software into the framework, and the generation of publication-quality
graphics for data visualization.

Data analysis for most experiments usually begins with a quality
assurance step, and we can use functionality from the flowQ package to
create an HTML report that highlights potential quality
issues. Assuming that the data has already been imported as the
flowSet object ``dat'', the following simple line of code produces the
quite complex output:

\begin{verbatim}
> library(flowCore)
> library(flowQ)
> qaReport(dat, c("qaProcess.timeline", "qaProcess.timeflow", 
                  "qaProcess.cellnumber"))
\end{verbatim}

The report is interactive and provides drilldown to more detailed
aspects of the analysis, starting from a concise overview
(Figure~\ref{flowQ}). The design of flowCore's data model allows for a
coherent treatment of all the samples, hence we are able to compare
features between individuals, or between groups of individuals, based
on the available annotation information.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{flowQ.jpg}
\caption{\label{flowQ}% 
  HTML quality assesment report generated by
  the flowQ package. Rows correspond to samples, columns to different
  quality checks.}
\end{figure}

Static gating for all samples in a high-throughput FCM experiment is
often not an option, since the measured variables tend to vary between
different treatments, over time or between different experiment
batches. Automated or data-driven gating tries to estimate the gating
regions from the underlying data, thus providing a fast objective
solution to the analysis of potentially very large and diverse data
sets \cite{lo2008agf}. One of the automated gating methods implemented
in flowCore is based on identifying areas of significant curvature in
a kernel density estimate of the data \citep{wand2008}. Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in a one or two-dimensional density landscape.

\begin{verbatim}
> cf <- curv2Filter("FL1-H", "FL3-H")
> fres <- filter(dat, cf)
\end{verbatim}

Kernel density estimation is a well-known problem in statistical
computing, and a lot of effort has been invested in the development of
good software to address it. The modular design of flowCore allows to
easily integrate these existing solutions into our framework. In this
example, we directly use R code from the feature package developed by
\cite{wand2008}. Instead of re-writing existing code, we are able to
include it via the well tested distribution mechanism provided by R's
software package system. This process is bi-directional, and all
functionality implemented in flowCore is available to other package
authors, as we have seen with the afore-mentioned flowClust package.

We can chose one of the many visualization option from the flowViz
package to plot the results of the recent filtering operation. A very
basic matrix of density plots is shown in Figure \ref{xyplot}, where
each panel in the matrix represents the data of one individual
patient.

\begin{verbatim}
> xyplot(`FL1-H` ~ `FL3-H` | SampleID, data=dat, filter=fres)
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics[width=0.75\textwidth]{xyplot}
\caption{\label{xyplot}%
Scatterplot matrix of a single flowSet with outlines of the gating
regions identified by an automated gating operation.}
\end{figure}

\section*{Discussion}
